Student Log-Data from a Randomized Evaluation of Educational Technology: A Causal Case Study

### 2020-01-06 paper.Rnw
working on comment 36

$\hat{\Delta}$ estimates the difference between the average outcome
for students in the entire treatment group, and the average outcome
for the treatment group were $H=0$ for all students.
That is
\begin{equation*}
\Delta=\EE[Y(1,H(1))-Y(1,0)|Z=1]
\end{equation*}
This can be re-written:
\begin{align*}
\begin{split}
\Delta&=\EE[Y(1,1)-Y(1,0)|Z=1,H=1]Pr(H=1)\\
&\qquad +\EE[Y(1,0)-Y(1,0)|Z=1,H=0]Pr(H=0)
\end{split}
\\[2ex]
&=\EE[Y(1,1)-Y(1,0)|Z=1,H=1]Pr(H=1)
&=Pr(H=1)TOT
\end{align*}
Since $Pr(H=1)=$\Sexpr{round(1-mixProp,2)},
$\hat{\Delta}=$\Sexpr{round(1-mixProp,2)}$TOT$.
Estimated thusly, the indirect effect is negative: \Sexpr{round(tott[1]*(1-mixProp),2)}, with
a standard error of \Sexpr{round(tott[2]*(1-mixProp),2)}.
This analysis suggests that the availability of hints
actually \emph{lowers} the treatment effect, that CTA1 works not
because of hints, but despite them.

@@

We incorporated the model for hint requests, (\ref{eq:rasch}), into
the larger principal stratification model, replacing $\eta(1)$ for
$\bar{h}(1)$.
Then, the principal stratification estimand is
\begin{equation*}
\tau(r)=\EE[Y(1)-Y(0)|\eta(1)=r]
\end{equation*}
A full treatment of this variant of principal stratification, including a
discussion of identification and estimation, may be found in
\citet{aoas}.

@@

First, students' proclivity to request hints is measured by the $\eta$
parameter in (\ref{eq:rasch}).
In the next level, $\eta(1)$ is modeled as a function of baseline
covariates $\bm{x}$:
\begin{equation}\label{eq:rasch2}
\eta(1)|\left(\bm{x}_i,\bm{\theta}\right) \sim
\mathcal{N}\left(\bm{x_i}\bm{\beta}^M+\epsilon^{Mt}_{t[i]}+\epsilon^{Ms}_{s[i]},
\sigma^M\right)
\end{equation}
where $\bm{\beta}^U$ is a vector of coefficients.
Since students were nested within teachers, who were nested within
schools, we included normally-distributed school ($\epsilon_{Us}$) and
teacher ($\epsilon_{Ut}$) random intercepts.
The covariates in the model, $\bm{x}_i$, were detailed in Table
\ref{tab:covariateBalance}; preliminary model checking suggested
including a quadratic term for pretest, which was added as a column of
$\bm{x}_i$.

We modeled students' posttest scores $Y$ as
conditionally normal:
\begin{equation}\label{eq:outcomeSubmodel}
 Y|\left(Z_i,\bm{x}_i,\bm{\theta},\eta(1)\right) \sim  \mathcal{N}\left(
\beta^Y_{0b[i]}+\bm{x}_i^T\bm{\beta}^Y+a\eta(1)+Z_i\tau(\eta(1))+\epsilon^{Yt}_{t[i]}+\epsilon^{Ys}_{s[i]},\sigma^Y_{Z[i]}\right)
\end{equation}
where $\beta^Y_{0b[i]}$ is a fixed effect for $i$'s randomization block, $\bm{\beta}^Y$ are the
covariate coefficients, and $\epsilon^{Yt}$, and
$\epsilon^{Ys}$ are normally-distributed teacher and school random
intercepts.
The residual variance $\sigma^Y$ varies with treatment assignment $Z$;
this captures measurement error in $Y$, treatment effect heterogeneity
that is not linearly related to
$\eta(1)$, and other between-student variation in $Y$ that is not predicted by
the mean model.

@@

Observational studies, and mediation analyses that build on them,
require ignorability assumptions such as Strong Ignorability or
Matched Ignorability.
These may be particularly problematic in our study, since students who
are more likely to request hints are also more likely to struggle with
the material, and hence score lower on the posttest.
We addressed this concern in three ways: first, we focused our attention
on the set of problems in the tutor in which students either requested
a hint or made an error, second, with propensity score matching, and
third with sensitivity analysis.
None of these approaches is bulletproof---they each require further
assumptions about how students request hints.
Eliminating problems in which students showed no signs of struggling
is likely to reduce the association between hint requests and
underlying academic ability, but not eliminate it.
For instance, stronger students may be better able to figure out a
difficult problem without access to hints or error feedback than
weaker students.
Matching depends both on the adequacy of the observed covariates to
capture baseline differences between $H=0$ and $H=1$ students, and the
sensitivity analyses we performed assumed that any un-measured
covariate would predict $H$ no better than pretest, or no better than
a covariate half as important as pretest.
These assumptions are impossible to verify from the data at hand;
instead, they require a keen understanding of student practice within
the tutor.

@@

In contrast, no ignorabilty assumptions are necessary for principal
stratification.
That said, an assumption similar to ignorability may play an important
role for interpretation.
Further, principal stratification estimators require a model for $Y(0)$
conditional on $\eta(1)$, which is unobserved.
On the other hand, model misspecification is, in principle,
detectable---an assumption of a particular family of models is not
``untestable.''
Still, the data will, in general, be unable to distinguish between certain
alternative models for $Y(0)$ conditional on $\eta(1)$, that may in
turn lead to qualitatively different conclusions.
In practice, analysts make the untestable assumption that $Y(0)$ conditional on $\eta(1)$ is
distributed similarly to $Y(1)$ conditional on $\eta(1)$.

@@

Our observational study found that hint requests have a negative effect: students who
requested hints more often than the \Sexpr{round((1-mixProp)*100)}th
percentile score lower on the posttest, after controlling for observed covariates---as
far as we can tell, the optimal strategy is to never request a hint.
However, this observed effect may be due to confounding.
Mediation analysis essentially interprets this result in terms of the
overall CTA1 treatment effect.
Specifically, our analysis showed that CTA1 affected students'
posttest scores in (at least) two opposite ways: by allowing hints, it
lowered their posttest scores, but its other mechanisms increased
their scores by an even greater amount.

>>>>

$\hat{\Delta}$ estimates the difference between the average outcome
for students in the entire treatment group, and the average outcome
for the treatment group were $H=0$ for all students.
That is
\begin{equation*}
\Delta=\EE[Y(1,H(1))-Y(1,0)|Z=1]
\end{equation*}
This can be re-written:
\begin{align*}
\begin{split}
\Delta&=\EE[Y(1,1)-Y(1,0)|Z=1,H=1]Pr(H=1)\\
&\qquad +\EE[Y(1,0)-Y(1,0)|Z=1,H=0]Pr(H=0)
\end{split}
\\[2ex]
&=\EE[Y(1,1)-Y(1,0)|Z=1,H=1]Pr(H=1)\\
&=Pr(H=1)TOT
\end{align*}
Since $Pr(H=1)=$\Sexpr{round(mixProp,2)},
$\hat{\Delta}=$\Sexpr{round(mixProp,2)}$TOT$.
Estimated thusly, the indirect effect is negative: \Sexpr{round(effects[['tot_eff']][['coefficients']]*(mixProp),2)}, with
a standard error of \Sexpr{round(tott[2]*(mixProp),2)}.
This analysis suggests that the availability of hints
actually \emph{lowers} the treatment effect, that CTA1 works not
because of hints, but despite them.

@@

We incorporated the model for hint requests, (\ref{eq:rasch}), into
the principal stratification model, replacing $\eta(1)$ for
$\bar{h}(1)$.
Then, the principal stratification estimand is
\begin{equation*}
\tau(r)=\EE[Y(1)-Y(0)|\eta(1)=r]
\end{equation*}
A full treatment of this variant of principal stratification, including a
discussion of identification and estimation, may be found in
\citet{aoas}.

@@

First, students' proclivity to request hints is measured by the $\eta$
parameter in (\ref{eq:rasch}).
In the next level, $\eta(1)$ is modeled as a function of baseline
covariates $\bm{x}$:
\begin{equation}\label{eq:rasch2}
\eta(1)|\left(\bm{x}_i,\bm{\theta}\right) \sim
\mathcal{N}\left(\bm{x_i}\bm{\beta}+\epsilon^{tch}_{t[i]}+\epsilon^{scl}_{s[i]},
\sigma\right)
\end{equation}
where $\bm{\beta}$ is a vector of coefficients.
Since students were nested within teachers, who were nested within
schools, we included normally-distributed school ($\epsilon^{scl}$) and
teacher ($\epsilon^{tch}$) random intercepts.
The covariates in the model, $\bm{x}_i$, were detailed in Table
\ref{tab:covariateBalance}; preliminary model checking suggested
including a quadratic term for pretest, which was added as a column of
$\bm{x}_i$.

We modeled students' posttest scores $Y$ as
conditionally normal:
\begin{equation}\label{eq:outcomeSubmodel}
 Y|\left(Z_i,\bm{x}_i,\bm{\theta},\eta(1)\right) \sim  \mathcal{N}\left(
\gamma_{0b[i]}+\bm{x}_i^T\bm{\gamma}+a\eta(1)+Z_i\tau(\eta(1))+\zeta^{tch}_{t[i]}+\zeta^{scl}_{s[i]},\omega_{Z[i]}\right)
\end{equation}
where $\bamma_{0b[i]}$ is a fixed effect for $i$'s randomization block, $\bm{\gamma}$ are the
covariate coefficients, and $\zeta^{tch}$, and
$\zeta^{scl}$ are normally-distributed teacher and school random
intercepts.
The residual variance $\omega$ varies with treatment assignment $Z$;
this captures measurement error in $Y$, treatment effect heterogeneity
that is not linearly related to
$\eta(1)$, and other between-student variation in $Y$ that is not predicted by
the mean model.

@@

Observational studies, and mediation analyses that build on them,
require ignorability assumptions such as Strong Ignorability or
Matched Ignorability.
These may be particularly problematic in our study, since students who
are more likely to request hints are also more likely to struggle with
the material, and hence score lower on the posttest.
We addressed this concern in three ways: first, we focused our attention
on the set of problems in the tutor in which students either requested
a hint or made an error, second, with propensity score matching, and
third with sensitivity analysis.
None of these approaches is bulletproof---they each require further
assumptions about how students request hints.
Eliminating problems in which students showed no signs of struggling
is likely to reduce the association between hint requests and
underlying academic ability, but not eliminate it.
For instance, stronger students may be better able to figure out a
difficult problem without access to hints or error feedback than
weaker students.
Matching depends both on the adequacy of the observed covariates to
capture baseline differences between $H=0$ and $H=1$ students, and the
sensitivity analyses we performed assumed that any un-measured
covariate would predict $H$ no better than pretest, or no better than
a covariate half as important as pretest.
These assumptions are impossible to verify from the data at hand;
instead, judging their plausibility requires a keen understanding of student practice within
the tutor.

@@

In contrast, no ignorabilty assumptions are necessary for principal
stratification.
% That said, an assumption similar to ignorability may play an important
% role for interpretation.
However, principal stratification estimators require modeling
assumptions---specifically, it requires a model for $Y(0)$
conditional on $\eta(1)$, which is unobserved.
This modeling assumption is testable, but only to an extent.
Model misspecification checks can compare a fitted model to observed
data, but

Certain departures from this assumption are detectable,
On the other hand, model misspecification is, in principle,
detectable---an assumption of a particular family of models is not
``untestable.''
Still, the data will, in general, be unable to distinguish between certain
alternative models for $Y(0)$ conditional on $\eta(1)$, that may in
turn lead to qualitatively different conclusions.
In practice, analysts make the untestable assumption that $Y(0)$ conditional on $\eta(1)$ is
distributed similarly to $Y(1)$ conditional on $\eta(1)$.

@@

Our observational study suggested that hint requests have a negative effect: students who
requested hints more often than the \Sexpr{round((1-mixProp)*100)}th
percentile score lower on the posttest, after controlling for observed covariates---as
far as we can tell, the optimal strategy is to never request a hint.
However, this observed effect may be due to confounding.
Mediation analysis essentially interprets this result in terms of the
overall CTA1 treatment effect.
Specifically, our analysis showed that CTA1 affected students'
posttest scores in (at least) two opposite ways: by allowing hints, it
lowered their posttest scores, but its other mechanisms increased
their scores by an even greater amount.
